`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models

### 2025-10-08 paper/paper.qmd
clean up name space dependencies blabla

---
title: ""
author: ""
format:
  pdf:
    latex-engine: pdflatex    # lualatex or xelatex or pdflatex
    include-before-body: title.tex
    include-in-header: 
      - file: preamble.tex 
editor: visual
execute:
  cache: true

abstract: | 
  Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations. Such models are typically fitted to a global database, e.g., to produce national-level estimates for all countries in the world or in some region. To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, such as data for one country. We refer to such models as local models: models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.

@@

```{r}
#| include: false
library(here)
library(tidyverse)
library(ggplot2)
library(haven)
library(stringr)
library(cmdstanr)
library(truncnorm) # for inits
library(bayesplot) # diagnostic plots
library(posterior) # else error for as_draws_list
library(tidybayes)
library(patchwork)
library(localhierarchy)
```

@@

```{r}
dat_nat |> select(level1,  mu_global, level1_eta, y)
```

@@

```{r}
global_fit$post_summ |> 
  filter(variable_no_index == "mu_sigma")
```

@@

```{r, message=FALSE, warning=FALSE}
res_global <- posterior_summary_hierparam(fit = global_fit, parname = "mu")
res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")
```

@@

The column `name` includes the country names, given by simple indicares in this simulation. The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:

```{r}
plots <- plot_posterior_summaries(
  res = res_global, 
  modelname1 = "global",
  hierarchy_select = "level1",
  areas_select = unique(res_local$level1$name)[1:4]
)

@@

```{r}
plots <- plot_posterior_summaries(
  res = res_local, 
  modelname1 = "local",
  res2 = res_global, 
  modelname2 = "global", 
  hierarchy_select = "level1",
  areas_select = unique(res_local$level1$name)[1:4]
)

@@

```{r}
plots <- plot_posterior_summaries(
  res = res_local, 
  modelname1 = "local",
  res2 = res_global, 
  modelname2 = "global"
)

@@

```{r, message=FALSE, warning=FALSE}
plots <- plot_mu_raw(fit = global_fit, parname = "mu")
```

@@

```{r}
dat_subnat <- dat_all$data
dat_subnat |> select(level1, level2, mu_global, level1_eta, level2_eta, y)
```

@@

```{r}
global_fit$post_summ |>
  filter(str_detect(variable_no_index, "sigma"))
global_fitsubnat_seq$post_summ |>
  filter(str_detect(variable_no_index, "sigma"))
```

@@

```{r, message=FALSE, warning=FALSE}
res_globalsubnat_seq <- posterior_summary_hierparam(
  fit = global_fitsubnat_seq,
  parname = "mu"
)

res_local_seq <- posterior_summary_hierparam(
  fit = fit_local_seq, 
  parname = "mu"
)
```

```{r}
plots <- plot_posterior_summaries(
  res = res_local_seq, 
  modelname1 = "local",
  res2 = res_globalsubnat_seq, modelname2 = "global",
  hierarchy_select = "level2",
  areas_select = unique(res_local_seq$level2$name)[1:30]
)

@@

| **Argument** | **Description** |
|----|----|
| **`survey_df`** | Tibble with survey data |
| **`y`** | Column name of outcome, defaults to **`y`**. |
| **`area`** | Column name of the area of each observation (used as geounit; iso or subnational region) |
| **`area_select`** | Area name to use for local run (e.g., iso code or subnational region name) |
| **`runstep`** | Type of run, defines which model fitting step to perform (see Details for options). |
| **`global_fit`** | Optional global fit object, used to obtain fixed values for some parameters in the current fit (see Details). |
| **`hierarchical_level`** | Vector specifying hierarchical structure used for mu. Should list elements in hierarchical order from `intercept` downward (see Details for examples). |
| **`add_subnational_hierarchy`** | Level that's added to the hierarchy for subnational analysis, defaults to `subnat`. |
| **`use_globalsubnat_fromnat`** | Logical; for local subnational runs, whether to use the global fit derived from national data (requires **`fit_globalsubnat_fromnat`** in **`global_fit`** if TRUE). |
| **`mu_isvector`** | Logical, TRUE if mu is a vector, defaults to FALSE |
| **`chains`** | Number of chains to run (sampling parameter) |
| **`iter_sampling`** | Number of posterior samples to draw (sampling parameter) |
| **`iter_warmup`** | Number of warmup iterations (sampling parameter) |
| **`compile_model`** | Boolean indicator of whether to compile the Stan model |
| **`force_recompile`** | Boolean indicator of whether to force recompilation of the Stan model |
| **`seed`** | Random seed |
| **`refresh`** | Number of iterations between progress updates |
| **`adapt_delta`** | Target acceptance rate for the No-U-Turn Sampler |
| **`max_treedepth`** | Maximum tree depth for the No-U-Turn Sampler |

>>>>

---
title: ""
author: ""
format:
  pdf:
    latex-engine: pdflatex    # lualatex or xelatex or pdflatex
    include-before-body: title.tex
    include-in-header: 
      - file: preamble.tex 
editor: visual
execute:
  cache: true
abstract: | 
  Bayesian hierarchical models are widely used in global health estimation, where data availability may vary across national or subnational populations. Such models are typically fitted to a global database, e.g., to produce national-level estimates for all countries in the world or in some region. To facilitate analysis at a local level, models are often desirable that are informed by global models but can be fitted to just a subset of the data, such as data for one country. We refer to such models as local models: models that are derived from global hierarchical models but adapted such that they can be fitted to the data from one population alone.

@@

```{r}
#| include: false
library(localhierarchy)
```

@@

```{r}
dat_nat |> dplyr::select(level1,  mu_global, level1_eta, y)
```

@@

```{r}
global_fit$post_summ |> 
  dplyr::filter(variable_no_index == "mu_sigma")
```

@@

```{r, message=FALSE, warning=FALSE}
res_global <- posterior_summary_hierparam_localhierarchy(
  fit = global_fit, parname = "mu")
res_local <- posterior_summary_hierparam_localhierarchy(
  fit = fit_local, parname = "mu")
```

@@

The column `name` includes the country names, given by simple indices in this simulation. The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:

```{r}
plots <- plot_posterior_summaries_localhierarchy(
  res = res_global, 
  modelname1 = "global",
  hierarchy_select = "level1",
  areas_select = unique(res_local$level1$name)[1:4]
)

@@

```{r}
plots <- plot_posterior_summaries_localhierarchy(
  res = res_local, 
  modelname1 = "local",
  res2 = res_global, 
  modelname2 = "global", 
  hierarchy_select = "level1",
  areas_select = unique(res_local$level1$name)[1:4]
)

@@

```{r}
plots <- plot_posterior_summaries_localhierarchy(
  res = res_local, 
  modelname1 = "local",
  res2 = res_global, 
  modelname2 = "global"
)

@@

```{r, message=FALSE, warning=FALSE}
plots <- plot_muraw_localhierarchy(fit = global_fit, parname = "mu")
```

@@

```{r}
dat_subnat <- dat_all$data
dat_subnat |> dplyr::select(level1, level2, mu_global, level1_eta, level2_eta, y)
```

@@

```{r}
global_fit$post_summ |>
  dplyr::filter(stringr::str_detect(variable_no_index, "sigma"))
global_fitsubnat_seq$post_summ |>
  dplyr::filter(stringr::str_detect(variable_no_index, "sigma"))
```

@@

```{r, message=FALSE, warning=FALSE}
res_globalsubnat_seq <- posterior_summary_hierparam_localhierarchy(
  fit = global_fitsubnat_seq,
  parname = "mu"
)

res_local_seq <- posterior_summary_hierparam_localhierarchy(
  fit = fit_local_seq, 
  parname = "mu"
)
```

```{r}
plots <- plot_posterior_summaries_localhierarchy(
  res = res_local_seq, 
  modelname1 = "local",
  res2 = res_globalsubnat_seq, modelname2 = "global",
  hierarchy_select = "level2",
  areas_select = unique(res_local_seq$level2$name)[1:30]
)

@@

| **Argument** | **Description** |
|------------------------------------|------------------------------------|
| **`survey_df`** | Tibble with survey data |
| **`y`** | Column name of outcome, defaults to **`y`**. |
| **`area`** | Column name of the area of each observation (used as geounit; iso or subnational region) |
| **`area_select`** | Area name to use for local run (e.g., iso code or subnational region name) |
| **`runstep`** | Type of run, defines which model fitting step to perform (see Details for options). |
| **`global_fit`** | Optional global fit object, used to obtain fixed values for some parameters in the current fit (see Details). |
| **`hierarchical_level`** | Vector specifying hierarchical structure used for mu. Should list elements in hierarchical order from `intercept` downward (see Details for examples). |
| **`add_subnational_hierarchy`** | Level that's added to the hierarchy for subnational analysis, defaults to `subnat`. |
| **`use_globalsubnat_fromnat`** | Logical; for local subnational runs, whether to use the global fit derived from national data (requires **`fit_globalsubnat_fromnat`** in **`global_fit`** if TRUE). |
| **`mu_isvector`** | Logical, TRUE if mu is a vector, defaults to FALSE |
| **`chains`** | Number of chains to run (sampling parameter) |
| **`iter_sampling`** | Number of posterior samples to draw (sampling parameter) |
| **`iter_warmup`** | Number of warmup iterations (sampling parameter) |
| **`compile_model`** | Boolean indicator of whether to compile the Stan model |
| **`force_recompile`** | Boolean indicator of whether to force recompilation of the Stan model |
| **`seed`** | Random seed |
| **`refresh`** | Number of iterations between progress updates |
| **`adapt_delta`** | Target acceptance rate for the No-U-Turn Sampler |
| **`max_treedepth`** | Maximum tree depth for the No-U-Turn Sampler |
